Dysregulation of Cerebral Hemoglobin Content in Malawian Children with Malaria

### 2021-07-08 figures.Rmd
adjusting figures based on new data format, still waiting for final DFA results to create logistic regression model

load("glm_model.Rdata")

@@

# read in data
df_hbox <- read.csv("Data/98 first visit records with cerebral oxygenation stats_trimmed.csv") %>% as_tibble()

# get rid of "Other"
df_hbox <- df_hbox %>% 
  filter(Status != "Other") %>% 
  mutate(Status = factor(Status, levels = c("HV", "UM", "CM")))

# identify duplicated rows (TM0003, TM2001)
df_hbox <- df_hbox %>% filter(duplicated(Subject.ID..NIAID.) == FALSE)

# Convert age to numeric
df_hbox <- df_hbox %>% 
  mutate(Age = gsub(" Years, ", "_", Age)) %>% 
  mutate(Age = gsub(" Months ", "", Age)) %>% 
  separate(Age, into = c("Years", "Months"), sep = "_") %>% 
  mutate(Years = as.numeric(Years)) %>% 
  mutate(Months = as.numeric(Months)/12) %>% 
  mutate(Age = Years + Months) %>% 
  dplyr::select(-c(Years, Months))

# Convert sex to factor
df_hbox <- df_hbox %>% mutate(Sex = as.factor(Sex))

# Trim for variables of interest (VOI)
df_hbox_trim <- df_hbox %>% 
  dplyr::select(c(Subject.ID..NIAID., Status, Glucose, Hematocrit, Lactate, Temperature,
                  Arginine.umol.L, Haptoglobin..mg.dl., Hemoglobin..uM., 
                  BP.Diastolic, BP.Systolic, O2.Sat, RR, HR,

                  cerebral.hbdeox, cerebral.hbdeox.alpha, cerebral.hbox, cerebral.hbox.alpha,
                  cerebral.sat, cerebral.sat.alpha, cerebral.thc, cerebral.thc.alpha)) %>%

  dplyr::rename("subject_id" = "Subject.ID..NIAID.", 
                "Arginine (umol/L)" = "Arginine.umol.L", "Haptoglobin (mg/dL)" = "Haptoglobin..mg.dl.", 
                "Hemoglobin (uM)" = "Hemoglobin..uM.", "Lactate (mmol/L)" = "Lactate", "Hematocrit (%)" = "Hematocrit",
                "Systolic BP" = "BP.Systolic", "Diastolic BP" = "BP.Diastolic", "O2 sat" = "O2.Sat", 
                "Respiration rate" = "RR", "Heart rate" = "HR",

                "[Hb]" = "cerebral.hbdeox", "[Hb]α" = "cerebral.hbdeox.alpha", 
                "[HbO2]" = "cerebral.hbox", "[HbO2]α" = "cerebral.hbox.alpha",  
                "HbO2%" = "cerebral.sat", "HbO2% α" = "cerebral.sat.alpha", 
                "[HbT]" = "cerebral.thc", "[HbT]α" = "cerebral.thc.alpha")

# Remove patients labeled as "UM" that instead meet the criteria for severe malaria

  # remove UM patient with lactate > 5 
  df_hbox_trim <- df_hbox_trim %>% dplyr::filter(!(Status == "UM" & `Lactate (mmol/L)` > 5))

  # All UM patients have hematocrit > 15%
  # df_hbox_trim %>% dplyr::filter(Status == "UM"  & `Hematocrit (%)` <= 15)

df_hbox_impute <- df_hbox_trim %>% 
  group_by(Status) %>% 
  mutate_at(3:ncol(.), f_impute) %>% 
  ungroup()

# pivot long
df_hbox_long <- df_hbox_trim %>% pivot_longer(cols = 3:ncol(.), names_to = "variable", values_to = "value")

@@

# number of subjects 
df_status <- df_hbox_impute %>% 
  count(Status) %>% 
  pivot_wider(1:2, names_from = "Status", values_from = "n") %>% 
  mutate(variable = "Number of subjects") %>% 
  dplyr::select(variable, everything()) %>% 
  mutate_if(is.numeric, as.character)

# Sex
df_sex <- df_hbox %>% 
  group_by(Status) %>% 
  count(Sex) %>% 
  filter(Sex != "") %>% 
  pivot_wider(names_from = "Sex", values_from = "n") %>% 
  mutate(perc_fem = round(Female/(Female + Male)*100, 2)) %>%

@@

  dplyr::filter(Subject.ID..NIAID. != "TM1018") %>%

@@

df_hbox_voi <- df_hbox_impute %>% 
  dplyr::select(c(Status, `[HbO2]`, `[HbO2]α`, `[Hb]`, `[Hb]α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
  dplyr::filter(`[Hb]α` < 3)

df_hbox_voi_long <- df_hbox_voi %>% 
  pivot_longer(2:9, names_to = "variable", values_to = "value") %>%
  mutate(variable = factor(variable, levels = c("[HbO2]", "[HbO2]α",   
                                                 "[Hb]", "[Hb]α", 
                                                "HbO2%", "HbO2% α", 
                                                "[HbT]", "[HbT]α")))

my_col <- c("CM" = "#00BFC4", "HV" = "#7CAE00", "UM" = "#F8766D")

df_hbox_voi_long %>% dplyr::filter(variable %in% c("[HbO2]", "[Hb]", "HbO2%", "[HbT]")) %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
  facet_wrap(~variable, scales = "free_y", nrow = 2, ncol = 2) +

@@

# [HbT]
kruskal.test(`[HbT]` ~ Status, data = df_hbox_voi)
dunnTest(`[HbT]` ~ Status, data = df_hbox_voi, method = "bh")

# HbO2%
kruskal.test(`HbO2%` ~ Status, data = df_hbox_voi)

@@

**Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status [HbT] levels were different between all three groups (Kruskal-Wallis adjusted p-value = 3.25x10^-6^; Dunn Test post-hoc adjusted p-values: CM vs HV = 1.66x10^-6^, CM vs UM = .032, HV vs UM = .033). Levels of cerebral O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = .07).

@@

df_hbox_voi <- df_hbox_impute %>% 
  dplyr::select(c(Status, `[HbO2]`, `[HbO2]α`, `[Hb]`, `[Hb]α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
  dplyr::filter(`[Hb]α` < 3)

df_hbox_voi_long <- df_hbox_voi %>% 
  pivot_longer(2:9, names_to = "variable", values_to = "value") %>%
  mutate(variable = factor(variable, levels = c("[HbO2]", "[HbO2]α",   
                                                 "[Hb]", "[Hb]α", 
                                                "HbO2%", "HbO2% α", 
                                                "[HbT]", "[HbT]α")))

my_col <- c("CM" = "#00BFC4", "HV" = "#7CAE00", "UM" = "#F8766D")

df_hbox_voi_long %>% dplyr::filter(variable %in% c("[HbO2]α", "[Hb]α", "HbO2% α", "[HbT]α")) %>%

@@

  geom_hline(yintercept = 0.4, alpha = 0.4, lty = 2) +
  geom_hline(yintercept = 0.6, alpha = 0.4, lty = 2) +
  geom_hline(yintercept = 1.0, alpha = 0.4, lty = 2) +
  geom_hline(yintercept = 1.5, alpha = 0.4, lty = 2) +

  facet_wrap(~variable, scales = "free_y", nrow = 2, ncol = 2) +

@@

# [HbT]α
kruskal.test(`[HbT]α` ~ Status, data = df_hbox_voi)
dunnTest(`[HbT]α` ~ Status, data = df_hbox_voi, method = "bh")

@@

**Figure 2** Shown are the alpha values associated with variables measured by the Oxiplex, plotted by patient status. Note: One outlier was removed from Deoxy Hb α (3.52).

@@

#### A. PCA
```{r fig3a}
# include variables of interest ([HbT], [HbT] alpha, and hematocrit), remove healthy volunteers from analysis
df_hbox_log_voi <- df_hbox_impute %>% 
  dplyr::select(c(2,4:5,19:22)) %>% 
  dplyr::filter(Status != "HV")

# convert tibble to df
df_hbox_voi_df <- df_hbox_log_voi %>% 
  dplyr::select(-Status) %>% 
  as.data.frame() 
rownames(df_hbox_voi_df) <- df_hbox_impute %>% dplyr::filter(Status != "HV") %>% .$subject_id

@@

# define groups
groups <- as.factor(df_hbox_log_voi$Status)

@@

### Figure 4: Relationships among [HbT], [HbT] alpha, and hematocrit
```{r fig4}

@@

  df %>%
    filter(Status == status) %>% 
    dplyr::select(-Status) %>% 
    cor(method = "spearman")

@@

# create df of potential regressors
df_hbox_reg <- df_hbox_impute %>% 
  dplyr::select(c(Status, `Hematocrit (%)`, `Lactate (mmol/L)`, `[HbO2]`, `[HbO2]α`, `[Hb]`, `[Hb]α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
  dplyr::filter(`[Hb]α` < 3)

m_cor_hv <- df_hbox_reg %>% f_calc_cor(status = "HV")

m_cor_um <- df_hbox_reg %>% f_calc_cor(status = "UM")

m_cor_cm <- df_hbox_reg %>% f_calc_cor(status = "CM")

@@

    mutate_if(is.factor, factor, levels = c("Lactate (mmol/L)", "Hematocrit (%)", "[HbO2]", "[Hb]", "HbO2%", "[HbT]", "[HbO2]α", "[Hb]α", "HbO2% α", "[HbT]α")) %>%

@@

**Figure S1.** Shown are the spearman correlations among variables measured by Oxiplex and their associated alpha values. Red indicates a positive correlation; blue indicates a negative correlation.

@@

# pivot longer and change factor levels
df_hbox_reg_long <- df_hbox_reg %>% 
  pivot_longer(2:11, names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = c("Lactate (mmol/L)", "Hematocrit (%)", "[HbO2]", "[HbO2]α", "[Hb]", "[Hb]α", "HbO2%", "HbO2% α" ,"[HbT]", "[HbT]α")))

# plot as individual regression curves
df_hbox_reg_long %>% mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>%

@@

**Figure S2.** Plotted are the individual logistic regression curves of individual variables tested for their discriminatory power between UM and CM. On the y-axis, 1 represents CM while 0 represents UM. A steeper curve indicates higher discriminatory power

>>>>

load("glm_model.Rdata")
load("results.Rdata")

@@

# read in data

  df_hbox <- read.csv("Data/master_datatable.csv") %>% 
    as_tibble() %>% 
    dplyr::select(-1) %>% 
    mutate(Status = factor(Status, levels = c("HV", "UM", "CM")))

# pivot long

  df_hbox_long <- df_hbox %>% pivot_longer(cols = 4:ncol(.), names_to = "variable", values_to = "value")

@@

# number of subjects 
df_status <- df_hbox %>% 
  count(Status) %>% 
  pivot_wider(1:2, names_from = "Status", values_from = "n") %>% 
  mutate(variable = "Number of subjects") %>% 
  dplyr::select(variable, everything()) %>% 
  mutate_if(is.numeric, as.character)

# Sex
df_sex <- df_hbox %>% 
  group_by(Status) %>% 
  count(Sex) %>% 
  dplyr::filter(Sex != "") %>% 
  pivot_wider(names_from = "Sex", values_from = "n") %>% 
  mutate(perc_fem = round(Female/(Female + Male)*100, 2)) %>%

@@

  # dplyr::filter(Subject.ID..NIAID. != "TM1018") %>%

@@

# extract variables of interest
voi <- df_results %>% colnames() %>% .[4:9]

# select from the long df
df_hbox_voi_long <- df_hbox_long %>% dplyr::filter(variable %in% voi)

# create color vector
my_col <- c("CM" = "#00BFC4", "HV" = "#7CAE00", "UM" = "#F8766D")

# plot
df_hbox_voi_long %>% dplyr::filter(!(grepl("alpha", variable))) %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
  facet_wrap(~variable, scales = "free_y") +

@@

# hemoglobin concentration
kruskal.test(avg_Hb_conc ~ Status, data = df_hbox)
dunnTest(avg_Hb_conc ~ Status, data = df_hbox, method = "bh")

# hemoglobin oxygen saturation
kruskal.test(avg_Hb_o2sat ~ Status, data = df_hbox)

@@

**Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status. Cerebral tissue hemoglobin concentration levels were different between cerebral malaria patients and the other two groups, but did not differ between uncomplicated malaria patients and healthy volunteers (Kruskal-Wallis adjusted p-value = 5.00x10^-7^; Dunn Test post-hoc adjusted p-values: CM vs HV = 3.22x10^-6^, CM vs UM = 2.76x10^-4^, HV vs UM = 0.87). Levels of cerebral hemoglobin O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = 0.28).

@@

df_hbox_voi_long %>% dplyr::filter(grepl("second_alpha", variable)) %>%

@@

  geom_hline(yintercept = 0.45, alpha = 0.4, lty = 2) +
  geom_hline(yintercept = 0.55, alpha = 0.4, lty = 2) +
  geom_hline(yintercept = 1.0, alpha = 0.4, lty = 2) +
  geom_hline(yintercept = 1.5, alpha = 0.4, lty = 2) +

  facet_wrap(~variable, scales = "free_y") +

@@

# hemoglobin concentration alpha
kruskal.test(Hb_conc_second_alpha ~ Status, data = df_hbox)
dunnTest(Hb_conc_second_alpha ~ Status, data = df_hbox, method = "bh")

# hemoglobin oxygen saturation alpha
kruskal.test(Hb_o2sat_second_alpha ~ Status, data = df_hbox)
dunnTest(Hb_o2sat_second_alpha ~ Status, data = df_hbox, method = "bh")

@@

**Figure 2.** Shown are the alpha values associated with the hemoglobin concentration and oxygen saturation signals measured by Oxiplex. Differences among cerebral tissue hemoglobin concentration alpha values were significant by Kruskal-Wallis (adjusted p-value = 1.57x10^-4^), and differences between all groups were significant by post-hoc Dunn Test (adjusted p-values: CM vs HV = 0.03, CM vs UM = 1.65x10^-4^, HV vs UM = 0.04). Cerebral hemoglobin O~2~ saturation alpha values differed between cerebral malaria patients and the other two patient groups (Kruskal-Wallis adjusted p-value = 2.71x10^-5^; Dunn Test post-hoc adjusted p-values: CM vs HV = 5.18x10^-4^, CM vs UM = 4.44x10^-4^, HV vs UM = 0.23).

@@

#### A. PCA
```{r fig3a}

# include variables of interest, remove healthy volunteers from analysis

df_hbox_voi <- df_hbox %>% dplyr::select(c("subject_id", "Status", "Hematocrit", "Lactate", "avg_Hb_conc", "Hb_conc_second_alpha", "avg_Hb_o2sat", "Hb_o2sat_second_alpha"))

  df_hbox_impute <- df_hbox_voi %>% 
    group_by(Status) %>% 
    mutate_at(3:ncol(.), f_impute) %>% 
    ungroup()

# remove healthy volunteers for logistic regression analysis
df_hbox_log_reg <- df_hbox_impute %>% dplyr::filter(Status != "HV")

# convert tibble to df
df_hbox_voi_df <- df_hbox_log_reg %>% 
  dplyr::select(-c(subject_id, Status)) %>% 
  as.data.frame()

# add rownames
rownames(df_hbox_voi_df) <- df_hbox_log_reg %>% .$subject_id

@@

# define groups
groups <- as.factor(df_hbox_log_reg$Status)

@@

### Figure 4: Relationships among hemoglobin concentration, hemoglobin concentration alpha, and hematocrit
```{r fig4}

@@

  df %>%
    dplyr::filter(Status == status) %>% 
    dplyr::select(-c(subject_id, Status)) %>% 
    cor(method = "spearman")

@@

# create run correlations of potential regressors
m_cor_hv <- df_hbox_impute %>% f_calc_cor(status = "HV")

m_cor_um <- df_hbox_impute %>% f_calc_cor(status = "UM")

m_cor_cm <- df_hbox_impute %>% f_calc_cor(status = "CM")

@@

    mutate_if(is.factor, factor, levels = c("Lactate", "Hematocrit", "avg_Hb_o2sat", "avg_Hb_conc", "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")) %>%

@@

**Figure S1.** Shown are the spearman correlations among hematocrit, lactate, hemoglobin concentration and its alpha, and hemoglobin oxygen saturation and its alpha. Red indicates a positive correlation; blue indicates a negative correlation.

@@

# pivot longer and change factor levels
df_hbox_reg_long <- df_hbox_impute %>% 
  mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>% 
  pivot_longer(3:ncol(.), names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = c("Lactate", "Hematocrit", "avg_Hb_o2sat", "avg_Hb_conc", "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")))

# plot as individual regression curves
df_hbox_reg_long %>%

@@

**Figure S2.** Plotted are the individual logistic regression curves of individual variables tested for their discriminatory power between UM and CM. On the y-axis, 1 represents CM while 0 represents UM. A steeper curve indicates higher discriminatory power.
